---
title: "Data visualization tutorial: exploring data and telling stories using ggplot2"
output:
  pdf_document:
    keep_tex: false
    latex_engine: pdflatex
    template: readable.tex
author:
  name: Peter Carbonetto
  affiliation: University of Chicago
date: July 29, 2020
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: single
graphics: yes
endnote: no
thanks: "This document is included as part of the Data Visualization tutorial packet for the BSD qBio Bootcamp, University of Chicago, 2020. **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: pcarbo@uchicago.edu. Thanks to John Novembre, Stephanie Palmer, Stefano Allesina and Matthew Stephens for their guidance."
abstract: "In this lesson, you will use ggplot2 to create effective
  visualizations of data. The ggplot2 package is a powerful plotting
  interface that extends the base plotting functions in R. You will
  learn that creating effective visualizations hinges on good data
  preparation. (In reality, good data preparation can take
  days or weeks, but we can still illustrate some useful data
  preparation practices.) The main difference with Advanced Computing
  1 is that we take a more in-depth look at ggplot2 and plotting
  strategies."
---

Motivation
==========

Why plot? One reason: to gain insight into your data. This is
"exploratory data visualization." Another reason: to tell a story
(e.g., for a scientific paper). Both require iteration and refinement
to get right. So you need to learn to create plots efficiently.

For this reason, we recommend the *programmatic approach* to data
visualization. This will allow you to:

1. Create an endless variety of plots.

2. Reuse code to quickly create and revise plots.

**ggplot2** is the the most powerful—and increasingly popular—approach
to creating plots in R.

```{r knitr-options, echo=FALSE}
knitr::opts_chunk$set(results = "hide",collapse = TRUE,
                      fig.height = 3,fig.width = 3,fig.align = "center")
```

Setup
=====

Download the tutorial materials to your computer, and make sure you
know where to find them.

Before starting the tutorial, I suggest quiting applications that are
not needed, and other "clutter", to reduce distractions.

Launch RStudio. It is best if you start with a fresh workspace; you
can refresh your environment by selecting **Session > Clear
Workspace** from the RStudio menu. Also, make sure your R working
directory is the same directory containing the tutorial materials; you
can run `getwd()` and `list.files()` to check this.

If you have not already done so, install these packages:

```{r install-pkgs, eval=FALSE}
install.packages("ggplot2")
install.packages("cowplot")
install.packages("ggrepel")
install.packages("htmlwidgets")
install.packages("plotly")
```

Hands-on exericse: Do smaller dogs live longer?
===============================================

In this tutorial, we will make use some data that was made available
by the authors of a 2008 *Genetics* article,
"Single-nucleotide-polymorphism-based association mapping of dog
stereotypes." These data are stored in a CSV file that was included
with the rest of the tutorial materials.

Import the data into R
----------------------

By now, you should know how to import data from a CSV file. This code
reads the data from a CSV file and creates a data frame, `dogs`. (Make
sure your working directory is set to the location of these data,
otherwise R will not be able to find the file.)

```{r read-data}
dogs <- read.csv("dogs.csv",stringsAsFactors = FALSE)
```

A first look at the data
------------------------

After loading data into R for the first time, I recommend getting a
basic understanding of the data frame and its contents. Here are some
simple commands I often use:

```{r inspect-data}
nrow(dogs)
ncol(dogs)
names(dogs)
head(dogs)
tail(dogs)
summary(dogs)
class(dogs$aod)
class(dogs$height)
class(dogs$weight)
class(dogs$shortcoat)
```

Note that "aod" stands for "age of death".

Let's take a closer look at the `shortcoat` column:

```{r inspect-shortcoat}
unique(dogs$shortcoat)
```

Feel free to explore the data further.

The often overlooked scatterplot
--------------------------------

In this tutorial, we will learn about ggplot2 through one of the most
basic data visualizations: the scatterplot.
 
The scatterplot is easily overlooked because it is so simple. But it
can be one of the most effective ways to visualize scientific
data. Occasionally, embellishments (e.g., varying colors, shapes,
sizes, labels) can produce stunning visualizations.

Typically, the purpose of the scatterplot is to investigate whether
there is an interesting—or surprising—relationship between two
(continuous) variables. But the scatterplot has many other uses,
particularly in exploratory data analysis. For example, the
scatterplot can highlight problems with the data, e.g., "outliers". It
is also useful for investigating the distribution of the data.

Although simple, we can explore the key features of ggplot2 via the
scatterplot, and, in fact, its simplicity is helpful in not
distracting us from our main goal, which is to learn about how to
create good plots with ggplot2.

In exploratory data analysis, it is usually helpful to have a focus.
Here, our aim is to investigate the anecdotal claim that dogs
representing breeds of small size such as Chihuahuas live longer than
those from larger-sized breeds such as Saint Bernards.

Let's begin by creating a scatterplot of weight vs. longevity using
the base R function, `plot`. (No packages needed here.)

```{r basic-plot, fig.height=4, fig.width=3.5}
plot(dogs$weight,dogs$aod)
```

*Question:* Based on this plot, how would you describe, qualitatively,
the relationship between weight and longevity?

Our first ggplot2 (with ugly code)
----------------------------------

Now let's recreate this same scatterplot using ggplot2. The benefits
of ggplot2 over `plot` will not be immediately clear.

```{r scatterplot-1}
library(ggplot2)
p1 <- ggplot(dogs,aes_string(x = "weight",y = "aod"))
p1 <- ggplot_add(geom_point(),p1)
```

Where is the plot? We need to call `print`, which tells R to draw the
plot to the screen:

```{r print-scatterplot}
print(p1)
```

This code is indeed quite cumbersome, and some simpler code can accomplish
the same thing:

```{r scatterplot-2a}
p1 <- ggplot(dogs,aes_string(x = "weight",y = "aod")) +
  geom_point()
```

```{r scatterplot-2b, echo=FALSE}
print(p1)
```

For the moment I want to focus on the "uglier" code because it
highlights better the key elements of a ggplot2 plot:

1. The first input is a data frame.

2. The second input is an "aesthetic mapping" that defines how columns
   are mapped to features of the plot (axes, shapes, colors, *etc*).

3. A "geom", short for "geometric object", specifies the type of
   plot. ggplot2 has an excellent on-line reference
   ([ggplot2.tidyverse.org][https://ggplot2.tidyverse.org]) explaining
   all the "geoms", from bar charts to contour plots, with code
   examples for each.

4. Rather than draw directly to the screen, ggplot2 creates a *ggplot
   object*, which can be drawn to the screen at any time by calling
   `print`.

The distinguishing feature of ggplot2—and what makes ggplot2 so
powerful—is that plots are created by *adding layers*. This layering
is what allows for an infinite variety of plots to be created with
ggplot2. Once you have mastered this layering, you will find that you
can use ggplot2 to create sophisticated, high-quality plots with
little effort. Further, the layering approach means that ggplot2 is
easily extendible, and many R packages have been developed to enhance
ggplot2. (We will use two of these packages, ggrepel and cowplot.)

In the remainder, we will explore this layering approach to plotting.

Some improvements
-----------------

Our plot can be improved. Here we add "layers" to adjust plot's
appearance, including:

+ Clearly label the axes.

+ Add a title.

+ Change the location of the axis tick marks.

+ Choose a more attractive theme (I'm partial to the cowplot
  theme).

Can you think of other ways the plot can be improved?

```{r scatterplot-improved, message=FALSE}
library(cowplot)
p1 <- ggplot(dogs,aes_string(x = "weight",y = "aod")) +
  geom_point() +
  scale_x_continuous(breaks = seq(0,160,40)) +
  labs(x = "body weight (lbs)",
       y = "longevity (years)",
       title = "longevity vs. size") +
  theme_cowplot()
```

```{r scatterplot-improved-2, echo=FALSE}
print(p1 +
  theme_cowplot(font_size = 12) +
  theme(plot.title = element_text(size = 12,face = "plain")))
```

Plot the best-fit line
----------------------

It has been estimated that an increase of 28 lbs in a dog's body
weight corresponds to about a 1-year drop in expected lifespan (with a
maximum lifespan of about 13 years). Let's see how well this estimate
agrees with the data by plotting the line that best fits these data
(the "least-squares" estimate).

```{r lm}
fit <- lm(aod ~ weight,dogs)
```

Add two "abline" layers to compare the estimates:

```{r scatterplot-with-abline}
p2 <- p1 +
  geom_abline(color = "dodgerblue",linetype = "dashed",
              intercept = coef(fit)[1],slope = coef(fit)[2]) +
  geom_abline(color = "darkorange",linetype = "dashed",
              intercept = 13,slope = -1/28)
```

```{r scatterplot-with-abline-2, echo=FALSE}
print(p2 +
  theme_cowplot(font_size = 12) +
  theme(plot.title = element_text(size = 12,face = "plain")))
```
Notice how easy it was to add layers to the existing plot. You can't
do this quite so easily with other plotting interfaces. Also notice
that this geom, like most geoms, has properties such as color and size
that can be adjusted.

Here I created a new plot object, `p2`, rather than overwrite the
existing one, `p1`. Now that I have both plots, I can decide later on
which one I want to keep.

How well do the two linear regressions agree?

Add the breed names
-------------------

It would be helpful if we could tell which breeds are being
plotted. Adding text to a ggplot is done in two steps: (1) map a column
to the "label" aesthetic, (2) add a `geom_text` or `geom_label`
geom.

Adding all the breed names will not work, however—there simply isn't
enough real estate on the plot to accommodate all the breed
names. Instead, a "smarter" function is `geom_text_repel` from the
ggrepel package, which adds the labels in a way that makes them more
readable, and only adds them to the plot when possible. Let's
appreciate again how simply this is accomplished.

```{r scatterplot-with-labels}
library(ggrepel)
p3 <- p2 + geom_text_repel(mapping = aes_string(label = "breed"),
                           size = 2.5,color = "gray")
```

```{r scatterplot-with-labels-2, fig.height=3.5, fig.width=3.5, echo=FALSE}
print(p3 + 
  theme_cowplot(font_size = 12) +
  theme(plot.title = element_text(size = 12,face = "plain")))
```

I chose to draw the labels in light gray so as to not drown out the
scatterplot. Feel free to adjust the size, color and other
properties.

Also notice that the labels are automatically redrawn as the plot is
resized—give it a try.

Which breeds are outliers from the linear trend?

A surprising subtlety with color
--------------------------------

Other aesthetics—color, size, shape, *etc*—can also be used enhance a
scatterplot and tell a good story. In the last chapter of our
scatterplot, we will visualize the `shortcoat` column with color, and
in the process discover a complication with adding color to a plot.

The `shortcoat` column, you may recall, had values of 0 or 1 (with a
few NAs).

```{r inspect-shortcoat-2}
dogs$shortcoat
```

In principle, varying the color of the points should be straightforward.

```{r scatterplot-with-color}
p4 <- ggplot(dogs,aes_string(x = "weight",y = "aod",label = "breed",
                             color = "shortcoat")) +
  geom_point() +
  theme_cowplot()
```

```{r scatterplot-with-color-2, fig.width=4, echo=FALSE}
print(p4 + theme_cowplot(font_size = 12))
```

It "worked" in the sense that we were able to create a plot with
color, but what we did was not completely correct. What is incorrect?

This isn't a problem with the data itself. Rather, it is a problem of
*data representation*: the discrete-valued, or *categorical* data are
stored as a *continuously-valued* variable (numeric). The key is to
change how the data are represented, and convert to a categorical
variable, which in R is called a "factor":

```{r shortcoat-factor}
dogs$shortcoat <- factor(dogs$shortcoat)
```

This illustrates the more general point that careful data
preparation is important for creating good plots.

```{r scatterplot-with-color-3}
p4 <- ggplot(dogs,aes_string(x = "weight",y = "aod",label = "breed",
                             color = "shortcoat")) +
  geom_point() +
  theme_cowplot()
```

```{r scatterplot-with-color-4, fig.width=4, echo=FALSE}
print(p4 + theme_cowplot(font_size = 12))
```

One the few complaints I have with ggplot2 is that the default colors
choices are poor. So you will often need to make adjustments to the
color scheme. One rule-of-thumb is that "warmer" colors (e.g., orange,
red) tends to draw the reader's attention. There are several good
resources on use of color in data visualization, and I will mention a
couple here: Color Brewer (https://colorbrewer2.org); and a short
article, "Color blindness", by Bang Wong in *Nature Methods*
(2011). For more extensive discussion, see "Fundamentals of data
visualization" by Claus Wilke. In our scatterplot, the best color
choice is less clear, so I will let you experiment with different
choices. To override the color defaults, add a "scale_color_manual"
layer, for example,

```{r scatterplot-with-better-color}
p5 <- p4 + scale_color_manual(values = c("firebrick","tomato"),
                              na.value = "black")
```

```{r scatterplot-with-better-color-2, fig.width=4, echo=FALSE}
print(p5 + theme_cowplot(font_size = 12))
```

There are several different ways to specify colors, e.g., by name or
RGB value. I prefer the named colors; to get the full list of named
colors in R, run `colors()`. Here, I chose two similar colors because
I didn't want `shortcoat` to be a major focus of the plot. Also,
notice that the color of NAs is controlled by a separate argument.

The function that controls the color of a discrete variable has an odd
name: `scale_color_manual`. This is because, in ggplot2, all methods
that control the mapping of variables to colors, shapes, sizes, axes,
*etc*, start with `scale_`. For example, plotting weight on the
square-root scale could help to better space out the breeds with
smaller weights, and can be done by setting the `trans` argument
(short for "transformation") in `scale_x_continuous`:

```{r scale-x-continuous, fig.keep="none"}
p5 + scale_x_continuous(trans = "sqrt")
```

```{r scale-x-continuous-2, fig.width=4, echo=FALSE}
print(p5 + scale_x_continuous(trans = "sqrt") + theme_cowplot(font_size = 12))
```

Let's add back in the breed names and the regression lines for our final
plot:

```{r scatterplot-final}
p5 <- p5 + 
  geom_text_repel(size = 2.5,color = "gray") +
  geom_abline(color = "dodgerblue",linetype = "dashed",
              intercept = coef(fit)[1],slope = coef(fit)[2]) +
  geom_abline(color = "darkorange",linetype = "dashed",
              intercept = 13,slope = -1/28)
```

```{r scatterplot-final-2, fig.height=3.5, fig.width=4.5, echo=FALSE}
print(p5 + theme_cowplot(font_size = 12))
```

Obviously, we only touched the surface of ggplot2. But once you are
comfortable with these basic elements, you will find that almost
everything else in ggplot2 is a variation of what we covered in this
lesson.

*Optional exercise:* Try varying shape instead of color. Use
`scale_shape_manual` to select the shapes. Run `plot(0:23,pch =
0:23)` to look up the R shape numbers.

*Optional exercise:* Try varying size of the points instead of their
color. What problem do you run into, and what is the solution?

Save your work
--------------

This is a good point to save our work in an image file that can be
shared with others.

```{r save-plot}
ggsave("dogs.pdf",p4,height = 4,width = 4.5)
ggsave("dogs.png",p4,height = 4,width = 4.5)
```

*Question:* How does the PDF and PNG differ? When should you save the
plot as a PDF, and when should you save as a PNG?

Programming challenge: Mapping the genetic basis of physiological and behavioral traits in outbred mice
=========================================================================================================

In this programming challenge, you will use simple visualizations to
gain insight into biological data.

You are working in a lab studying studying the genetics of
physiological and behavioral traits in mice. The lab has just
completed a large study of mice from an outbred mouse population,
"CFW" ("Carworth Farms White"). The aim of the study is to identify
genetic contributors to variation in behaviour and musculoskeletal
traits.

**Note:** These challenges are roughly ordered in increasing level of
complexity. Do not be discouraged if you have difficulty completing
every one.

Collaboration strategy
----------------------

Before diving into the problems, first agree on a collaboration
strategy with your teammates. Important aspects include communication
and co-ordination practices, and setting goals and deadlines. How will
your team collaborate on code, and share solutions? (Consider online
resources such as Etherpad or the UofC-hosted Google Drive.) The aim
is not just to complete the challenges, but also to do
collaboratively; all team members should be included, and should
have the opportunity to contribute and learn from each other.

Instructions
------------

+ Locate the files for this exercise on your computer (see "Materials"
  below).

+ Make sure your R working directory is set to the same directory
  containing the tutorial materials; use `getwd()` to check this.

+ Some of the programming challenges require uploading an image file
  containing a plot. Use `ggsave` to save your plot as a file. Any
  standard image format (e.g., PDF, PNG) is acceptable.

+ No additional R packages are needed beyond what you used in the
  hands-on exercises above.

Materials
---------

+ **pheno.csv:** CSV file containing physiological and behavioral
phenotype data on 1,219 male mice from the CFW outbred mouse
stock. Data are from [Parker *et al*, 2016][cfw-paper]. Use
**readpheno.R** to read the phenotype data from the CSV file into a
data frame. After filtering out some of the samples, this script
should create a new data frame, `pheno`, containing phenotype data on
1,092 samples (rows).

+ **hmdp.csv:** CSV file containing bone-mineral density measurements
taken in 878 male mice from the Hybrid Mouse Diversity Panel
(HMDP). Data are from [Farber *et al*, 2011][hmdp-paper]. To load the
data into your R environment, run this code:

    ```{r read-hdmp-data}
    hmdp <- read.csv("hmdp.csv", stringsAsFactors = FALSE)
    ```

    This will create a data frame, `hmdp`, containing BMD data on
    878 mice (rows).

+ **gwscan.csv:** CSV file containing results of a "genome-wide scan"
for abnormal BMD. Association *p*-values were computed using
GEMMA 0.96. To read the results of the genome-wide scan, run
the following code:

    ```{r read-gwscan-results}
    gwscan <- read.csv("gwscan.csv", stringsAsFactors = FALSE)
    gwscan <- transform(gwscan, chr = factor(chr, 1:19))
    ```

    This will create a data frame, `gwscan`. Each row of the data
    frame is a genetic variant (a single nucleotide polymorphism, or
    "SNP"). The columns are chromosome ("chr"), base-pair position on
    the chromosome ("pos"), and the *p*-value for a test of
    association between variant genotype and trait value
    ("abnormalBMD"). The value stored in the "abnormalBMD" column is
    $-\log_{10}(P)$, where $P$ is the *p*-value.

+ **geno_rs29477109.csv:** CSV file containing estimated genotypes at
  one SNP (rs29477109) for 1,038 CFW mice. Use the following code to
  read the genotype data into your R environment:

    ```{r read-genotypes}
    geno <- read.csv("geno_rs29477109.csv", stringsAsFactors = FALSE)
    geno <- transform(geno, id = as.character(id))
    ```

    This will create a new data frame, `geno`, with 1,038 rows
    (samples). The genotypes are encoded as "dosages"---that is, the
    expected number of times the alternative allele is observed in the
    genotype. This will be an integer (0, 1, 2), or a real number
    between 0 and 2 when there is some uncertainty in the estimate of
    the genotype. For this SNP, the reference allele is T and the
    alternative allele is C. Therefore, dosages 0, 1 and 2 correspond
    to genotypes TT, CT and CC, respectively (genotypes CT and TC are
    equivalent).

+ **wtccc.png:** Example genome-wide scan ("Manhattan plot") from
Fig. 4 of the [WTCCC paper][wtccc-paper]. The *p*-values highlighted
in green show the regions of the human genome most strongly associated
with Crohn's disease risk.

A couple tips
-------------

+ Some "geoms" you may find useful: `geom_point`, `geom_histogram`,
  `geom_boxplot`.

+ In some cases it may be useful to convert to a *factor*.

Part A: Exploratory analysis of muscle development and conditioned fear data
----------------------------------------------------------------------------

Your first task is to create plots to explore the data.

1. A basic initial step in an exploratory analysis is to visualize the
distribution of the data. It is often convenient if the distribution
is normal, or "bell shaped".

    + Visualize the empirical distribution of tibialis anterior (TA)
      muscle weight (column "TA") with a histogram. Units are mg.
	  *Hint:* Try using function `geom_histogram`.

    + Is the distribution of TA weight roughly normal? Are there mice
      with unusually large or unusually small values ("outliers")? If
      so, how many "outliers" are there? (Unusually small or large
      values can lead to misleading results in some statistical
      tests.)

2. It is also important to understand relationships among measured
quantities. For example, the development of the tibia bone (column
"tibia") could influence TA muscle weight. Create a scatterplot
(`geom_point`) to visualize the relationship between TA weight and
tibia length. (Tibia length units are mm.) Based on this plot, what
can you say about the relationship between TA weight and tibia length?
Quantify this relationship by fitting a linear model, before and after
removing the outlying TA values. (*Hint*: Use the `lm` and `summary`
functions. See also the "r.squared" return value in
`help(summary.lm)`.)

3. The "AvToneD3" column contains data collected from a behavioral
test called the "Conditioned Fear" test.

    + Visualize the empirical distribution of AvToneD3 ("freezing to
      cue") with a histogram. Is the distribution of AvToneD3
      approximately normal?

    + Freezing to cue is a proportion (a number between 0 and 1). A
      common way to obtain a more "normal" quantity is to
      transform it using the "logit" function[^1]. Visualize the
      empirical distribution of the logit-transformed phenotype. Is
      the transformed phenotype more "bell shaped"? After the
      transformation, do you observe unusually small or unusually
      large values?

    + A common concern with behavioral tests is that the testing
      devices can lead to measurement error. It is especially a
      concern when multiple devices are used, as the devices can give
      slightly different measurements, even after careful calibration.
      Create a plot to visualize the relationship between
      (transformed) freezing to cue and the device used ("FCbox"
      column). *Hint:* Try a boxplot (`geom_boxplot`). Based on this
      plot, does the apparatus used affect these behavioral test
      measurements?

[^1]: R code: `logit <- function(x) log((x + 0.001) / (1 - x + 0.001))`

Part B: Exploratory analysis of bone-mineral density data
---------------------------------------------------------

Now you will examine data on bone-mineral density (BMD) in mice. This
is a trait that is important for studying human diseases such as
osteoporosis (units are $\mathrm{mg}/\mathrm{cm}^2$).

+ Plot the distribution of BMD in CFW mice (see column "BMD"). What
is most notable about the distribution?

+ Compare these data against BMD measurements taken in a "reference"
mouse population, the Hybrid Mouse Diversity Panel. To compare, create
two histograms, and draw them one on top of the other. What difference
do you observe in the BMD distributions? For a correct comparison, you
will need to account for: (1) BMD in CFW mice was measured in the
femurs of male mice only; (2) BMD in HMDP mice was recorded in
$\mathrm{g}/\mathrm{cm}^2$. *Hints:* Functions `xlim` and `labs` from
the ggplot2 package, and `plot_grid` from the cowplot package, might
be useful for creating the plots. The `binwidth` argument in
`geom_histogram` may also be useful.

Part C: Mapping the genetic basis of osteopetrotic bones
--------------------------------------------------------

A binary trait, "abnormal BMD", was defined that signals whether an
individual mouse had "abnormal", or osteopetrotic, bones. It takes a
value of 1 when BMD falls on the "long tail" of the distribution (BMD
greater than 90 $\mathrm{mg}/\mathrm{cm}^2$), otherwise zero.

GEMMA was used to carry out a "genome-wide association study" (GWAS)
for this trait; that is, support for association with abnormal BMD was
evaluated at 79,824 genetic variants (single nucleotide polymorphisms,
or "SNPs") on chromosomes 1--19. At each SNP, a *p*-value quantifies
the support for an association with abnormal BMD.

1. Your first task is to get an overview of the association results by
creating a "Manhattan plot". Follow as closely as possible the
provided prototype, **wtccc.png**, which shows a genome-wide scan for
Crohn's disease. (Don't worry about highlighting the strongest
*p*-values in green.) *Hints:* Replicating some elements of this
plot may be more challenging than others, so start with a simple plot,
and try to improve on it. Recall the adage that creating plots
requires relatively little effort *provided the data are in the right
form*---consider adding appropriate columns to the `gwscan` data
frame. Functions from the ggplot2 package that you may find useful for
this exercise include `geom_point`, `scale_color_manual` and
`scale_x_continuous`.

    + In your plot, you should observe that the most strongly
      associated SNPs cluster closely together in small regions of the
      genome. This is common---it is due to a genetic phenomenon known
      as linkage disequilibrium (LD). It is a consequence of low
      recombination rates between markers in small populations. How
      many SNPs have "strong" statistical support for association with
      abnormal BMD, specifically with a $-\log_{10}\mbox{{\em
      p}-value} > 6$? How many distinct regions of the genome are
      strongly associated with abnormal BMD at this *p*-value
      threshold?

    + What *p*-value does a $-\log_{10}\mbox{{\em p}-value}$ of 6
      correspond to?

    + Using your plot, identify the "distinct region" (this is called
      a "quantitative trait locus", or QTL) with the strongest
      association signal. What is, roughly, the size of the QTL in
      Megabases (Mb) if we define the QTL by base-pair positions of
      the SNPs with $-\log_{10}\mbox{{\em p}-value} > 6$? Using the
      [UCSC Genome Browser][ucsc-browser], get a rough count of the
      number of genes that are transcribed in this region. Within this
      QTL, [Parker *et al*, 2016][cfw-paper] identified *Col1a1* as a
      candidate BMD gene. Was this gene one of the genes included in
      your count? *Hint:* All SNP positions are based on NCBI Mouse
      Genome Assembly 38 (mm10, December 2011).

2. Your next task is to visualize the relationship between genotype
and phenotype. From the genome-wide scan of abnormal BMD, you should
find that rs29477109 is the SNP most strongly associated with
abnormal BMD. Here you will look closely at the relationship between
BMD and the genotype at this SNP. In developing your visualization,
consider that:

     + The samples listed in the phenotype and genotype tables are not
       the same. So you will need to align the two tables to properly
       show analyze the relationship. *Hint:* Function `match` could be
       useful for this.

     + The genotypes, stored in file **geno_rs29477109.csv**, are
       encoded as "dosages" (numbers between 0 and 2). You could start
       with a scatterplot of BMD vs. dosage. But ultimately it is more
       effective if the genotypes (CC, CT and TT) are
       plotted instead. *Hints:* In effect, what you need to do is
       convert from a continuous variable (dosage) to a discrete
       variable (genotype). One approach is to create a `factor`
       column from the "dosage" column. For dosages that are not
       exactly 0, 1 or 2, you could simply round to the nearest whole
       number. A boxplot is recommended; see function `geom_boxplot`.

    Based on your plot, how would describe (in plain language) the
    relationship between the genotype and BMD?

Notes
=====

Code summary
------------

This is the minimal code reproducing the final scatterplot:

```{r recap}
library(ggplot2)
library(ggrepel)
library(cowplot)
dogs <- read.csv("dogs.csv",stringsAsFactors = FALSE)
dogs$shortcoat <- factor(dogs$shortcoat)
fit <- lm(aod ~ weight,dogs)
p5 <- ggplot(dogs,aes_string(x = "weight",y = "aod",label = "breed",
             color = "shortcoat")) +
  geom_point() +
  geom_text_repel(size = 2.5,color = "gray") +
  geom_abline(color = "dodgerblue",linetype = "dashed",
              intercept = coef(fit)[1],slope = coef(fit)[2]) +
  geom_abline(color = "darkorange",linetype = "dashed",
              intercept = 13,slope = -1/28) +
  scale_color_manual(values = c("firebrick","tomato"),
                     na.value = "black") +
  theme_cowplot()
```

An interactive plot
-------------------

Here is an example of a plot that allows the data to be explored
*interactively*:

```{r scatterplot-plotly, message=FALSE, warning=FALSE}
library(plotly)
library(htmlwidgets)
p <- plot_ly(data = dogs,type = "scatter",mode = "markers",
             x = ~weight,y = ~aod,color = ~shortcoat,
             text = ~sprintf("%s\n (%0.1f lbs, %0.1f years)",breed,weight,aod),
             marker = list(line = list(color = "white",width = 1),size = 9),
             hoverinfo = "text",width = 600,height = 500)
p <- layout(p,
            xaxis = list(title = "weight (lbs)"),
            yaxis = list(title = "longevity (years)"),
            hoverlabel = list(bgcolor = "white",bordercolor = "black"))
saveWidget(p,"dogs.html",selfcontained = TRUE)
```

[Plotly][plotly] is a powerful package for creating interactive plots,
with an interface similar to ggplot2. 

Useful online resources
-----------------------

+ [ggplot2 reference][ggplot2], where you will also find a ggplot2
cheat sheet. (This cheat sheet is also included in the tutorial
packet.)

+ [Fundamentals of Data Visualization][wilke] by Claus Wilke.

Sources
-------

+ The dogs breeds data were downloaded from the
  [*Genetics* journal website][jones-2008].

+ The CFW phenotype and genotype data were downloaded from [Data
  Dryad][cfw-data].

License
-------

Except where otherwise noted, all instructional material in this
repository is made available under the
[Creative Commons Attribution license (CC BY 4.0)][cc-by-human]. And,
except where otherwise noted, the source code included in this
repository are made available under the [OSI][osi]-approved
[MIT license][mit-license]. For more details, see the `LICENSE.md`
file included in the tutorial packet.

[ggplot2]: http://ggplot2.tidyverse.org
[plotly]: https://plot.ly/r
[wilke]: https://serialmentor.com/dataviz
[cfw-paper]: https://doi.org/10.1038/ng.3609
[cfw-data]: https://doi.org/10.5061/dryad.2rs41
[hmdp-paper]: https://doi.org/10.1371/journal.pgen.1002038
[wtccc-paper]: https://doi.org/10.1038/nature05911
[ucsc-browser]: http://genome.ucsc.edu
[cc-by-human]: https://creativecommons.org/licenses/by/4.0
[mit-license]: http://opensource.org/licenses/mit-license.html
[osi]: http://opensource.org
[jones-2008]: https://doi.org/doi:10.1534/genetics.108.087866
