---
title: "Data visualization tutorial: exploring data and telling stories using ggplot2"
output:
  pdf_document:
    keep_tex: false
    latex_engine: pdflatex
    template: latex/readable.tex
author:
  name: Peter Carbonetto
  affiliation: University of Chicago
date: August 5, 2018
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: single
graphics: yes
endnote: no
thanks: "This document is included as part of the Data Visualization tutorial packet for the BSD qBio Bootcamp, MBL, 2018. **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: pcarbo@uchicago.edu. Thanks to John Novembre and Matthew Stephens for their support and guidance."
abstract: "In this tutorial, we will use ggplot2 to create effective visualizations of biological data. The ggplot2 package is a powerful set of plotting functions that extend the base plotting functions in R. In this lesson, we will also see that creating effective visualizations in R *hinges on good data preparation.* In reality, good data preparation in the lab can take days or weeks, but here we can still illustrate some useful data preparation practices. The main difference with Advanced Computing 1 is that we take a more careful look at ggplot2 and strategies for data visualization."
---

# Hands-on exercise: "Rising Dough, Rising Neighbourhoods"

```{r knitr-options, echo=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,
                      message = FALSE,warning = FALSE)
```

## The scenario

You have just begun a summer internship at the
[Mansueto Institute for Urban Innovation][mansueto]. The institute has
formed a partnership with the City of Chicago to develop new ways of
quickly gaining insight into local economic activity throughout the
city.

A computer scientist working at the institute recently has been
exploring ways to measure economic health of neighbourhoods by
aggregating publicly available data on pizza restaurants throughout
the city. The city is excited about the potential for this project,
and would like to present a report to the Mayor. Unfortunately, the
computer scientist left a few weeks ago to start a research team at
Amazon's new Headquarters in the West Loop. (*Disclaimer:* In case you
haven't figured this out already, much of this story is fictional.)
So, as the new student intern, you have been tasked with building on
the computer scientist's work to put together a report.

The good news is that this computer scientist was diligent about
keeping track of the code and data files she used in her analyses. She
even went so far as to provide detailed comments in her code
explaining what the code does. (This is perhaps one of the less
realistic parts of this scenario.) Your advisor has provided you with
the R code and data files that were used to generate the plots that
will go in the report. Therefore, although your only experience in R
is from a statistics course you took during your undergrad, your
initial sense of dread has turned into cautious optimism.

## Instructions

+ Make sure you have downloaded the tutorial packet.

+ Locate the files for this exercise on your computer (see "Materials"
  below).

+ To run the code, you will need to have the following R packages
  installed on your computer: **ggplot2**, **cowplot** and **readr**.

+ Open the the R source file, **pizzaplots.R**, in RStudio, or in your
  favourite editor (e.g., emacs).

+ Make sure your R working directory is the same directory containing
  the tutorial materials; use `getwd()` to check this.

+ Follow the instructor for additional instructions (see also the
  slides PDF included with the tutorial packet).

## Materials

+ **pizzaplots.R:** R code you will use to reproduce the plots.

+ **Food_Inspections.csv.gz**: Compressed text file in CSV Format
containing the food inspection data that were downloaded from the
[Chicago Data Portal][chicago-data-portal].

## Follow-up programming challenges

1. Our first plot was a bar chart showing an estimate of the number of
new pizza restaurants per year in Chicago. Sometimes lines are more
effective than bars for showing trends.

    + If you wanted to show the trend as a line plot, how would you
      modify the plotting code in **pizzaplots.R** to do this? *Hint:*
      `geom_line` might be useful.

    + What change do you need to make to the `counts` data frame
      so that counts can be plotted as lines?

    + Which do you find more effective, the line plot or the bar
      chart? Identify one benefit of one plot over the other.

    + To submit your response, you will need to upload a file
      containing your final plot. Use `ggsave` to save your plot as a
      file.

2. Your advisor has asked you to colour the bars in the bar chart so
that they align with the colours used in the map. 

    + What change do you need to make to the `counts` data frame so
      that the years can be mapped to colors? *Hint:* The code used
      for creating the line plot may be useful here, too.

    + Submit the *combined* plot (bar chart & map). When submitting
      your plot, save it as a PDF using `ggsave`.

    + Why did your advisor request a PDF? Why is a PNG file less
      suitable for a final report or publication?

# Main programming challenge: "Mapping the genetic basis of physiological and behavioral traits in outbred mice"

In this programming challenge, you will use simple visualizations to
gain insight into biological data.

You have finished your summer internship at the Mansueto Institute,
and you are now embarking on your first graduate research project in a
lab studying studying the genetics of physiological and behavioral
traits in mice. The lab has just completed a large study of mice from
an outbred mouse population known as "CFW" (short for "Carworth Farms
White", the names of the scientists who bred the first CFW mice). The
ultimate aim of the study is to identify genetic contributors to
variation in behaviour and musculoskeletal traits.

*Note:* These challenges are roughly ordered in increasing level of
complexity. Do not be discouraged if you have difficulty completing
all the exercises. Also, do not hesitate to ask the instructors for
advice if you get stuck.

## Instructions

+ Locate the files for this exercise on your computer (see "Materials"
  below).

+ Make sure your R working directory is the same directory containing
  the tutorial materials; use `getwd()` to check this.

+ Follow the instructor for additional guidance (see also the slides
  included in the tutorial packet).

+ Some of the programming challenges require uploading an image file
  containing a plot. You can use `ggsave` to save your plot as a file;
  any standard image format is acceptable.

+ No additional R packages are needed beyond what you used in the
  hands-on exercise above.

## Materials

+ **pheno.csv:** Text file containing physiological and behavioral
phenotype data on 1,219 male mice from the CFW outbred mouse
stock. The data are stored in comma-delimited (CSV) format, with one
sample per line. Data are from [Parker *et al*, 2016][cfw-paper]. Use
**readpheno.R** to read the phenotype data from the CSV file into a
data frame. After discarding some of the samples, this script will
create a data frame, `pheno`, containing phenotype data on 1,092
samples, with one row per sample.

+ **hmdp.csv:** Text file in CSV format containing bone-mineral
density measurements taken in 878 male mice from the Hybrid Mouse
Diversity Panel (HMDP). Data are from
[Farber *et al*, 2011][hmdp-paper]. To load the data into your R
environment, run the following code.

    ```{r read-hdmp-data}
    hmdp <- read.csv("hmdp.csv",stringsAsFactors = FALSE)
    ```

    This will create a data frame, `hmdp`, containing BMD data on
    878 mice, with one mouse per row.

+ **gwscan.csv:** Text file in CSV format containing results of a
"genome-wide scan" for abnormal BMD. Association *p*-values were
computed using [GEMMA][gemma] 0.96. To read the results of the
genome-wide scan, run the following code:

    ```{r read-gwscan-results}
    gwscan <- read.csv("gwscan.csv",stringsAsFactors = FALSE)
    gwscan <- transform(gwscan,chr = factor(chr,1:19))
    ```

    This will create a data frame, `gwscan`. Each row of the data
    frame is a single genetic variant in the mouse genome (a single
    nucleotide polymorphism, or "SNP"). The columns give the
    chromosome ("chr"), base-pair position on the chromosome ("pos"),
    and the *p*-value for a test of association between variant
    genotype and trait value ("abnormalBMD"). The value stored in the
    abnormalBMD column is $-\log_{10}(P)$, where $P$ is the
    association  test *p*-value.

+ **geno_rs29477109.csv:** Text file in CSV format containing
  estimated genotypes at one SNP (rs29477109) for 1,038 CFW mice. Use
  the following code to read the genotype data into your R
  environment:

    ```{r read-genotypes}
    geno <- read.csv("geno_rs29477109.csv",stringsAsFactors = FALSE)
    geno <- transform(geno,id = as.character(id))
    ```

    This will create a new data frame, `geno`, with 1,038 rows, one
    for each sample (mouse). The genotypes are encoded as "dosages";
    specifically, the expected number of times the alternative allele
    is observed in the genotype. This will either be an integer (0, 1
    or 2), or a real number between 0 and 2 when there is some
    uncertainty in the estimate of the genotype. In this case, the
    reference allele is T and the alternative allele is C. Therefore,
    dosages 0, 1 and 2 correspond to genotypes TT, CT and CC,
    respectively (note genotypes CT and TC are equivalent).

+ **wtccc.png:** Example genome-wide scan ("Manhattan plot") taken
from Fig. 4 of the [WTCCC paper][wtccc-paper]. The *p*-values
highlighted in green show the regions of the human genome most
strongly associated with Crohn's disease risk.

## Part A: Exploratory analyses of muscle development and conditioned fear data

Your first task is to create plots to explore of some of the phenotype
data collected for the CFW study.

1. A basic initial step in an exploratory analysis is to visualize the
empirical distribution of the data. For several reasons (e.g., to
ensure validity of statistical tests used), it is convenient if the
empirical distribution is normal, or "bell shaped".

    + Visualize the empirical distribution of tibialis anterior (TA)
      muscle weight (column "TA") with a histogram. Units are mg.
	  *Hint:* Try function `geom_histogram`.

    + Is the distribution of TA weight roughly normal? Are there mice
      with unusually large or unusually small TA muscles (*i.e.*,
      "outliers")?  If so, how many "outliers" are there? It is
      sometimes important to identify outliers, since unusually small
      or large values can lead to misleading results in standard
      statistical tests.

2. It is also often important to understand the relationships among
the measured quantities to be analyzed. For example, the development
of the tibia bone (column "tibia") could influence TA muscle
weight. Create a scatterplot (`geom_point`) to visualize the
relationship between TA weight and tibia length. Units of tibia length
are mm. Based on this plot, what can you say about the relationship
between TA weight and tibia length? Also, quantify this relationship
by fitting a linear model, before and after removing the outlying TA
values. *Hint*: Use the `lm` and `summary` functions for this. If you
are unsure how to quantify the relationship, see the description of
the "r.squared" output in `help(summary.lm)`.

3. The "AvToneD3" column contains data collected from a behavioral
test called the "Conditioned Fear" test. Specifically, AvToneD3 is the
average proportion of time freezing on the third day of testing during
the presentation of tones (the conditioned stimulus). The shorthand
for this behavioral phenotype is "freezing to cue".

    + Visualize the empirical distribution of freezing to cue with a
      histogram. Is the distribution of AvToneD3 approximately normal?

    + Freezing to cue is a proportion (a number between 0 and 1). A
      common way to obtain a more normal-behaving proportion is to
      transform it using the "logit" function[^1]. Visualize the
      empirical distribution of the logit-transformed phenotype. Is
      the transformed phenotype more "bell shaped"? After the
      transformation, do you observe unusually small or unusually
      large values?

    + A common concern with behavioral tests is that the devices used
      in the tests can lead to measurement error. It is especially a
      concern when multiple devices are used, as the devices can give
      slightly different measurements, even after careful calibration.
      Create a plot to visualize the relationship between freezing to
      cue (the transformed version) and the device used (column
      "FCbox"). *Hint:* Try creating a boxplot using `geom_boxplot`.
      Based on this plot, would you say that the apparatus used
      affected the measurements in this behavioral test?

[^1]: An R implementation of the logit function:
`logit <- function(x) log((x + 0.001)/(1 - x + 0.001))`

## Part B: Exploratory analyses of bone-mineral density data

In this part, you will examine data on bone-mineral density in
mice. This is a trait that is important for studying human diseases
such as osteoporosis. The units of BMD are
$\mathrm{mg}/\mathrm{cm}^2$. (Strictly speaking, this is not a
density---it is "areal" BMD, which is easier to measure, and is
considered a good approximation to the true BMD.)

+ Plot the distribution of BMD in CFW mice (see column "BMD"). What
feature stands out from the histogram?

+ To investigate whether this feature is particular to the CFW mouse
population, compare these BMD data against BMD measurements taken in a
"reference" mouse population. As reference, we will use the Hybrid
Mouse Diversity Panel. To provide a direct visual comparison, create
two histograms, and draw them one on top of the other. What difference
do you observe in the BMD distributions? Note that BMD in the CFW was
measured in the femurs of male mice. Further, note that BMD in the
HMDP data set was recorded in $\mathrm{g}/\mathrm{cm}^2$. *Tips:*
Functions `xlim` and `labs` from the ggplot2 package, and `plot_grid`
from the cowplot package, might be useful for creating the plots. The
`binwidth` argument in `geom_histogram` may also be useful.

## Part C: Mapping the genetic basis of osteopetrotic bones

Based on the exploratory analyses of BMD in CFW and HMDP mice, we
defined a binary trait, "abnormal BMD", that signals whether an
individual mouse had abnormal, or osteopetrotic, bones. It takes a
value of 1 when BMD falls on the "long tail" of the observed
distribution (BMD greater than 90 $\mathrm{mg}/\mathrm{cm}^2$), and 0
otherwise.

We used [GEMMA][gemma] to carry out a "genome-wide association study"
(GWAS) for this trait; that is, we estimated support for association
between abnormal BMD and 79,824 genetic variants (single nucleotide
polymorphisms, or "SNPs") on chromosomes 1--19. At each SNP, we
computed a *p*-value to assess the support for association with
abnormal BMD.

1. After running the *p*-value computations, the next step in any GWAS
is to get an overview of the association results. Your task here is to
create a "Manhattan plot" summarizing the results. Follow as closely
as possible the provided prototype, **wtccc.png**, which shows a
genome-wide scan for Crohn's disease. This prototype plot is from an
influential paper that set many of the standards for conducting
genome-wide association studies. (Note you do not need to worry about
highlighting the strongest *p*-values in green since the threshold for
choosing the "strongest" *p*-values is not clearly defined in this
study.) *A few tips:* Replicating some elements of this plot may be
more challenging than others, so start with a simple plot, and try to
improve on it. Recall the adage that creating plots requires
relatively little effort *provided the data are in the right
form*---consider adding appropriate columns to the `gwscan`
data frame. Functions from the ggplot2 package that you may find
useful for this exercise include `geom_point`, `scale_color_manual`
and `scale_x_continuous`.

    + In your plot, you should observe that the most strongly
      associated SNPs cluster closely together in small regions of the
      genome. This is a common situation, and is due to a genetic
      phenomenon known as linkage disequilibrium (LD). It arises as a
      natural consequence of low recombination rates between markers
      in small populations. How many SNPs have "strong" statistical
      support for association with abnormal BMD, specifically with a
      $-\log_{10}\mbox{{\em p}-value} > 6$? How many distinct regions
      of the genome are strongly associated with abnormal BMD at this
      *p*-value threshold?

    + What *p*-value does a $-\log_{10}\mbox{{\em p}-value}$ of 6
      correspond to?

    + Using your plot, identify the "quantitative trait locus" (QTL)
      with the strongest association signal (this is the region where
      the strongest associations cluster). What is, roughly, the size
      of the QTL in Megabases (Mb), if we define the QTL by the
      base-pair positions of the SNPs with $-\log_{10}\mbox{{\em
      p}-value} > 6$? Using the [UCSC Genome Browser][ucsc-browser],
      get a rough count of the number of genes that are transcribed in
      this region. Within this QTL, [Parker *et al*, 2016][cfw-paper]
      identified *Col1a1* as a compelling candidate for being the
      causal BMD gene. Was this gene one of the genes included in your
      count? *Note:* All SNP positions are based on Mouse Genome
      Assembly 38 from the NCBI database (mm10, December 2011).

2. In this last exercise, your task is to visualize the relationship
between genotype and phenotype. From the genome-wide scan of abnormal
BMD, we identified rs29477109 as the SNP most strongly associated with
abnormal BMD. Here you will look closely at the relationship between
BMD and the genotype at this SNP. In developing your visualization,
consider the following:

     + The samples listed in the phenotype and genotype tables are not
       the same. So you will need to align the two tables to properly
       show analyze the relationship. *Hint:* Function `match` is
       useful for this.

     + The genotypes, stored in file **geno_rs29477109.csv**, are
       encoded as "dosages" (numbers between 0 and 2). You could start
       with a scatterplot of BMD vs. dosage. But ultimately it is more
       directly interpretable if the genotypes (CC, CT and TT) are
       plotted instead. *Hint:* In effect, what you need to do is
       convert from a continuous variable (dosage) to a discrete
       variable (genotype). One approach is to create a new `factor`
       column from the "dosage" column. For dosages that are not
       exactly 0, 1 or 2, you could simply round to the nearest whole
       number. A boxplot is recommended, which you can create with
       `geom_boxplot`.

    Based on your plot, how would describe in plain language the
    relationship between the genotype and BMD?

# Notes

## Useful online resources

+ [ggplot2 reference][ggplot2], where you will also find a ggplot2
cheat sheet. (This cheat sheet is also included in the tutorial
packet, and you may have seen it in a previous tutorial.)

+ [Fundamentals of Data Visualization][wilke] by Claus Wilke. If you
are interested, I've also generated a PDF version of this book from
Claus's source code. You can access the PDF [here][wilke-pdf] on the
[UChicago Box][box]. This is a book I wish I had when I started my
Ph.D!

## License

Except where otherwise noted, all instructional material in this
repository is made available under the
[Creative Commons Attribution license (CC BY 4.0)][cc-by-human]. And,
except where otherwise noted, the source code included in this
repository are made available under the [OSI][osi]-approved
[MIT license][mit-license]. For more details, see the `LICENSE.md`
file included in the tutorial packet.

## Session info

```{r, echo=FALSE}
library(readr)
library(ggplot2)
library(cowplot)
```

This next code chunk gives information about the computing
environment, including the version of R and the R packages, that were
used to test the tutorial examples.

```{r}
sessionInfo()
```

## Other notes

+ To generate this PDF, run `rmarkdown::render("handout.Rmd")` in R.

+ The Latex template used to generate the PDF from R Markdown was
modified from a [template][svmiller-repo] created by
[Steven Miller][svmiller].

+ The CFW phenotype and genotype data were downloaded from the [Data
  Dryad repository][cfw-data].

+ The City of Chicago food inspection data were downloaded from the
[Chicago Data Portal][chicago-data-portal]. Specifically, the data
were downloaded in CSV format from [here][food-inspection-data] on
August 3, 2018.

+ The Chicago community map was downloaded from
[here][chicago-community-map].

+ For background on the using fast food data to assess the economy,
[read this][bbc-story].

[mansueto]: https://miurban.uchicago.edu
[box]: https://uchicago.box.com
[ggplot2]: http://ggplot2.tidyverse.org
[wilke]: https://serialmentor.com/dataviz
[wilke-pdf]: https://uchicago.box.com/v/wilke
[svmiller]: https://github.com/svmiller
[svmiller-repo]: https://github.com/svmiller/svm-r-markdown-templates
[bbc-story]: http://www.bbc.com/future/story/20170202-how-fast-food-reveals-secrets-of-the-economy
[chicago-data-portal]: https://data.cityofchicago.org
[food-inspection-data]: https://data.cityofchicago.org/Health-Human-Services/Food-Inspections/4ijn-s7e5
[chicago-community-map]: https://home.chicagopolice.org/community/community-map
[cfw-paper]: https://doi.org/10.1038/ng.3609
[cfw-data]: https://doi.org/10.5061/dryad.2rs41
[hmdp-paper]: https://doi.org/10.1371/journal.pgen.1002038
[wtccc-paper]: https://doi.org/10.1038/nature05911
[gemma]: https://github.com/genetics-statistics/GEMMA
[ucsc-browser]: http://genome.ucsc.edu
[cc-by-human]: https://creativecommons.org/licenses/by/4.0
[cc-by-legal]: https://creativecommons.org/licenses/by/4.0/legalcode
[mit-license]: http://opensource.org/licenses/mit-license.html
[osi]: http://opensource.org
